Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-24 numerical-zeus.tex
Editing section 4.4 (zeus)

As discussed in the section describing the Piecewise Parabolic Method (Section~\ref{sec.hydro.ppm}), operator-split methods break the solution of the hydrodynamic equations into parts, with each part representing a single term in the equations.  Each part is evaluated successively using the results preceding it.  In this method, in addition to operator-splitting the expansion terms (i.e. those terms in eqs.~(\ref{eq:mass}) -- (\ref{eq:total_energy}) that depend on $\dot{a}$), we divide the remaining terms into \emph{source} and \emph{transport} steps.  The terms to be solved in the source step are those on the right-hand side of eqs.~(\ref{eq:mass}) -- (\ref{eq:total_energy}), while the transport terms are on the left-hand side of these equations and are responsible for the advection of mass, momentum and energy across the grid.

@@

As in the original \zeus\ paper, the source terms are added in three steps. First we add the pressure and gravity forces:
\begin{equation}
v_j^{n+a}  =  v_j^n - \frac{\Delta t}{\Delta x_j} \frac{p^n_j - p^n_{j-1}} {(\rho^n_j + \rho^n_{j-1})/2}.
\end{equation}
Partial updates are denoted by the $n+a$, $n+b$ notation.  We show updates in one dimension as the extension to the multi-dimensional case is straightforward (note that all dimensions are carried out for each substep before progressing to the next substep). We then add the artificial viscosity:
\begin{eqnarray}
v_j^{n+b} & = & v_j^{n+a} - \frac{\Delta t}{\Delta x_j} 
                             \frac{q^{n+a}_j - q^{n+a}_{j-1}} {(\rho^n_j + \rho^n_{j-1})/2} \\
e_j^{n+b} & = & e_j^n - \frac{\Delta t}{\Delta x_j} q^{n+a}_j (v^{n+a}_{j+1} - v^{n+a}_{j}).
\end{eqnarray}
The artificial viscosity coefficient $q_j$ is given by:
\begin{equation}
q_j = \left\{ \begin{array}{ll}
              Q_{\rm AV} \rho_j (v_{j+1} - v_j)^2 \quad & \rm{if} (v_{j+1} - v_j) < 0 \\
               0 & \rm{otherwise}
               \end{array} \right.
\end{equation}
where $Q_{\rm AV}$ is a constant with a typical value of 2. We refer the interested reader to \citet{Stone92a} and \citet{1994ApJ...429..434A} for more details.  We also include the option (turned off by default) of adding a linear artificial viscosity as suggested in the \zeus\ paper for stagnant flow regions.  This is given by
\begin{equation}
q_{{\rm lin},j} = Q_{\rm LIN} \rho c_j (v_{j+1} - v_{j})
\end{equation}
where $c_j^2 = \gamma p/\rho$ is the adiabatic sounds speed.

@@

Next, we examine the transport step, which is conservative.  Once again, we dimensionally split the equations and present only the one-dimensional version.  The finite difference equations actually solved are:
\begin{equation}
\rho_j^{n+d} = \rho_j^{n} - \frac{\Delta t}{\Delta x} (v^{n+c}_{j+1/2} \rho^{*}_{j+1/2} - v^{n+c}_{j-1/2} \rho^{*}_{j-1/2} )
\end{equation}
Here $\rho^*_j$ is the correctly upwinded value of $\rho$ evaluated at the cell-face corresponding to $v_j$, making $\rho^*_j v_j$ the mass flux at the cell boundary and guaranteeing mass conservation.   This requires interpolating each cell-centered quantity to the cell edge.  As recommended in \citet{Stone92a}, we use the second-order van Leer scheme, which uses piecewise linear functions.  These are given by equations (48) and (49) of \citet{Stone92a}.  The transport steps for the other variables are similar.  Note that we advect the specific energy and specific momenta using the mass flux, as dictated by the principle of consistent transport.  This requires appropriate averaging for the momenta in the perpendicular directions as outlined in equations (57)-(72) of \citet{Stone92a}.

>>>>

As discussed in the section describing the Piecewise Parabolic Method (Section~\ref{sec.hydro.ppm}), operator-split methods break the solution of the hydrodynamic equations into parts, with each part representing a single term in the equations.  Each part is evaluated successively using the results preceding it.  In this method, in addition to operator-splitting the expansion terms (i.e., those terms in Eqs.~(\ref{eq:mass}) -- (\ref{eq:total_energy}) that depend on $\dot{a}$), we divide the remaining terms into \emph{source} and \emph{transport} steps.  The terms to be solved in the source step are those on the right-hand side of eqs.~(\ref{eq:mass}) -- (\ref{eq:total_energy}), while the transport terms are on the left-hand side of these equations and are responsible for the advection of mass, momentum and energy across the grid.

@@

As in the original \zeus\ paper, the source terms are added in three steps. First we add the pressure and gravity forces:
\begin{equation}
v_j^{n+a}  =  v_j^n - \frac{\Delta t}{\Delta x_j} \frac{p^n_j - p^n_{j-1}} {(\rho^n_j + \rho^n_{j-1})/2}.
\end{equation}
Partial updates are denoted by the $n+a$, $n+b$ notation.  We show updates in one dimension as the extension to the multi-dimensional case is straightforward (note that all dimensions are carried out for each substep before progressing to the next substep). We then add the artificial viscosity:
\begin{eqnarray}
v_j^{n+b} & = & v_j^{n+a} - \frac{\Delta t}{\Delta x_j} 
                             \frac{q^{n+a}_j - q^{n+a}_{j-1}} {(\rho^n_j + \rho^n_{j-1})/2} \\
e_j^{n+b} & = & e_j^n - \frac{\Delta t}{\Delta x_j} q^{n+a}_j (v^{n+a}_{j+1} - v^{n+a}_{j}).
\end{eqnarray}
The artificial viscosity coefficient $q_j$ is given by:
\begin{equation}
q_j = \left\{ \begin{array}{ll}
              Q_{\rm AV} \rho_j (v_{j+1} - v_j)^2 \quad & \rm{if}~(v_{j+1} - v_j) < 0 \\
               0 & \rm{otherwise}
               \end{array} \right.
\end{equation}
where $Q_{\rm AV}$ is a constant with a typical value of 2. We refer the interested reader to \citet{Stone92a} and \citet{1994ApJ...429..434A} for more details.  We also include the option (turned off by default) of adding a linear artificial viscosity as suggested in the \zeus\ paper for stagnant flow regions.  This is given by
\begin{equation}
q_{{\rm lin},j} = Q_{\rm LIN} \rho c_j (v_{j+1} - v_{j})
\end{equation}
where $c_j^2 = \gamma p/\rho$ is the adiabatic sounds speed.

@@

We next examine the transport step, which is conservative.  Once again, we dimensionally split the equations and present only the one-dimensional version.  The finite difference equations actually solved are:
\begin{equation}
\rho_j^{n+d} = \rho_j^{n} - \frac{\Delta t}{\Delta x} (v^{n+c}_{j+1/2} \rho^{*}_{j+1/2} - v^{n+c}_{j-1/2} \rho^{*}_{j-1/2} )
\end{equation}
Here $\rho^*_j$ is the correctly upwinded value of $\rho$ evaluated at the cell-face corresponding to $v_j$, making $\rho^*_j v_j$ the mass flux at the cell boundary and guaranteeing mass conservation.   This requires interpolating each cell-centered quantity to the cell edge.  As recommended in \citet{Stone92a}, we use the second-order van Leer scheme, which uses piecewise linear functions.  These are given by Equations (48) and (49) of \citet{Stone92a}.  The transport steps for the other variables are similar.  Note that we advect the specific energy and specific momenta using the mass flux, as dictated by the principle of consistent transport.  This requires appropriate averaging for the momenta in the perpendicular directions as outlined in equations (57)-(72) of \citet{Stone92a}.
